---
title: "Exercise Performance"
author: "R. Baltimore"
date: "Saturday, July 25, 2015"
output: html_document
---

##Overview
This assignment was completed as a requirement for the Coursera Data Science Specialization 'Pracitical Machine Learning' Course. The aim of this project is to predict the manner in which weight-lifting exercises are being done, given personal activity data gathered from accelerometers on the belt, forearm, arm and dumbell of six (6) participants.Exercises were performed correctly and incorrectly in 5 different ways 


##Libraries
The following libraries are used in processing and model building:
```{r echo=TRUE, result="hide"}
library(caret)
library(randomForest)
```



##Data
Documentation for the data used for this project can be accessed from the website: [http://groupware.les.inf.puc-rio.br/har](http://groupware.les.inf.puc-rio.br/har):

The training and testing datasets are downloaded and loaded into R and the training set data examined
```{r echo=TRUE, results="hide"}
urlTrain <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
urlTest <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
#download.file(url=urlTrain, destfile='training.csv')
#download.file(url=urlTest, destfile='testing.csv')
training <- read.csv('training.csv')
testing <- read.csv('testing.csv')

str(training)
```

The data is cleaned to eliminate variables which contain mostly NA values, have little to no variance, and are not useful to our prediction process.
```{r echo=TRUE, result="hide"}
navals <- apply(training, 2, function(x) {sum(is.na(x))})
training <- training[,which(navals==0)]

nearzero <- nearZeroVar(training)
training <- training[, -nearzero]

nearzero2 <- nearZeroVar(testing)
testing <- testing[, -nearzero2]

training <- training[, -(1:5)]
testing <-testing[, -(1:5)]

```


The training set is partitioned into sets for training and cross-validation and the training partition is examined
```{r echo=TRUE, result="hide"}
set.seed(525)
inTrain <- createDataPartition(training$classe, p=0.6,list=FALSE)
partTrain <- training[inTrain, ]
partTest <- training[-inTrain, ]
str(partTrain)
```

##Developing Algorithm
The training partition now consists of 11776 observations of 54 variables. Training is done using the random forest algorithm on the training partition partTrain, while cross-validation is done on the testing partition partTest. 
```{r echo=TRUE}
modfit <- randomForest(classe~., data = partTrain, method = "class")

predictCV <- predict(modfit, partTest, type="class")
confusionMatrix(predictCV, partTest$classe)
```

The algorithm has quite a high accuracy at 0.9967 while the expected out-of-sample error is 0.0033. Hence, we are expecting than a very high number, if not all, of the testing set data will be classified correctly.

##Conclusion
We now use our model to predict the intial testing set.
```{r echo=TRUE}
predictTest <- predict(modfit, testing, type="class")
predictTest

```


